hatchR: A toolset to predict hatch and emergence phenology in wild fishes

### 2025-03-06 inst/manuscript/paper.Rmd
incoporate ms comments by @pfrater

Morgan M. Sparks^1,4^, Eli A. Felts^2^, Allison G. Swartz^3^, Bryan M. Maitland^1^

@@

^4^Corresponding Author: Morgan M. Sparks, Morgan.Sparks\@usda.gov

@@

The effective value approach has since been widely applied to Salmonids, for which aquaculture-derived parameterizations were readily available. For example, Pacific Salmon (*Oncorhynchus* spp.) models developed by @beacham1990 have been applied across various species and populations [@adelfio2024; @adelfio2019; @kaylor2021], while Bull Trout (*Salvelinus confluentus*) models from @mcphail1979 were extended by @austin2019. Despite its growing adoption, applications of the effective value model remain largely confined to Salmonids, likely due to the availability of existing parameterizations and the commercial and recreational importance of these species.

@@

To maximize accessibility, **hatchR** is available in two formats. The first is a R package, **hatchR**, distributed via The Comprehensive R Archive Network (CRAN), providing advanced customization and automation for analyzing multiple variables, such as phenology type, spawn timing, or thermal regimes. Comprehensive documentation is available on the **hatchR** website (<https://bmait101.github.io/hatchR/>). The second is a Shiny-based web application [@chang2024], offering a graphical user interface for those unfamiliar with R, balancing ease of use with much of the R package's core functionality. Below, we provide and overview of **hatchR** and present case studies demonstrating its application in research and management.

@@

$$
\mathrm{E_i} = \frac{1}{\exp(\log_e(a) - \log_e(T_i - b)}
$$

@@

**hatchR** assumes input data consists if at least two required columns: a date column indicating the date (and optionally time) of each temperature measurement, and a temperature column providing the corresponding temperature measurement (in °C). Other columns may be present, but columns names should not include spaces. Data should follow the format outlined in Table 1.

@@

We demonstrate the utility three functions: `summarize_temp()`, `plot_check_temp()`, and `check_continuous()`--using a simulated year-long data set (`year_sim`). This dataset contains temperature readings taken every thirty minutes, and its structure (dimensions and first six rows) is shown below.

@@

We demonstrate how the `fit_model()` function may be used to create custom parameterizations for species beyond the Salmonids included in `model_table`. To showcase its utility, we provide parameterizations for three warm-water species: Smallmouth Bass (*Micropterus dolomieu*) [@webster1948], Channel Catfish (*Ictalurus punctatus*) [@small2001], and Lake Sturgeon (*Acipenser fulvescens*) [@smith2005]. These species were selected due to their common use in aquaculture and sport fisheries, illustrating the broad applicability of the effective values approach. For concision, we present parameterization for Smallmouth Bass here, while the full implementation details for all species are available in the `paper.Rmd` on the GitHub project repository.

@@

To demonstrate this case study, we use the **hatchR** graphical user interface available at  <https://elifelts.shinyapps.io/hatchR_demo/>. Users begin by uploading their temperature dataset through the `Import Data` window, selecting their file, specifying the appropriate temperature and date columns, and providing the date format (e.g., year-month-day or day-month-year). For this example, we use the `crooked_river` dataset, included in the Shiny app as a demo dataset. It can also be accessed directly at [https://github.com/bmait101/hatchR/blob/master/extdata/crooked_river.csv/](#0){.uri}.

@@

Following model selection, **hatchR** outputs results in two key locations: the `Phenology Summaries` tab, which provides a table with predictions for each spawn date, and the `Timeline Plot` tab, which shows the corresponding visualization of development timing. Both the prediction table and plot can be downloaded directly from their respective tabs.

The process is demonstrated in the included ***supplementary video file???***, and a more detailed interface walk through is available on the **hatchR** Shiny website.

@@

We use the `idaho` dataset from @isaak2018, which includes four years of temperature data (2011-2014) for these sites. To identify putative Bull Trout Spawning locations, we apply a filtering criterion based on mean August temperature, as outlined in @isaak2015, selecting only sites with mean August temperature \</= 13 °C, a known thermal threshold for Bull Trout spawning suitability. The filtering process reduced the dataset to 139 sites potential spawning sites.

@@

With **hatchR**, we present a software ecosystem that bridges the analytical gap in predicting developmental phenology for wild fishes. It establishes a formal framework for applying effective value models from user-provided parameterizations. The software is available in two formats: 1) A fully customizable R package, ideal for complex and repetitive analyses and 2) a graphical user interface for ease of use, designed for tasks that may only need to be run once or a few times.

@@

Finally, while hatchR was designed specifically for fishes, it has potential applications for other poikilothermic organisms, such as reptiles, amphibians, and invertebrates, where developmental rates similarly follow a power law relationship with temperature. Extending the effective value framework to these taxa could provide valuable insights into their life history timing under variable environmental conditions.

@@

Importantly, hatchR extends the effective value framework developed by @sparks2019 into a generalizable tool that can be applied to any fish species or population, provided that appropriate source data are available—data that can easily be collected in aquaculture settings. We present foundational applications of **hatchR**, with additional user-friendly cases and implementation guides available on the software's website. The software is designed for both applied and basic research, allowing users to engage with it either through a programmatic R environment or via a user-friendly Shiny app. We expect that the examples provided here represent only a fraction of **hatchR**'s potential applications and encourage the user community to explore and expand upon this framework for their own research and management needs.

>>>>

Morgan M. Sparks^1,5^, Bryan M. Maitland^1^, Eli A. Felts^2^, Allison G. Swartz^3^, Paul Frater^4^

@@

^4^Wisconsin Department of Natural Resources, Madison, WI, USA

^5^Corresponding Author: Morgan M. Sparks, Morgan.Sparks\@usda.gov

@@

The effective value approach has since been widely applied to Salmonids, for which aquaculture-derived parameterizations are readily available. For example, Pacific Salmon (*Oncorhynchus* spp.) models developed by @beacham1990 have been applied across various species and populations [@adelfio2024; @adelfio2019; @kaylor2021], while Bull Trout (*Salvelinus confluentus*) models from @mcphail1979 were extended by @austin2019. Despite its growing adoption, applications of the effective value model remain largely confined to Salmonids, likely due to the availability of existing parameterizations and the commercial and recreational importance of these species.

@@

To maximize accessibility, **hatchR** is available in two formats. The first is an R package, **hatchR**, distributed via The Comprehensive R Archive Network (CRAN), providing advanced customization and automation for analyzing multiple variables, such as phenology type, spawn timing, or thermal regimes. Comprehensive documentation is available on the **hatchR** website (<https://bmait101.github.io/hatchR/>). The second format is a Shiny-based web application [@chang2024], offering a graphical user interface for those unfamiliar with R, balancing ease of use with much of the R package's core functionality. Below, we provide an overview of **hatchR** and present case studies demonstrating its application in research and management.

@@

$$
\mathrm{E_i} = \frac{1}{\exp(\log_e(a) - \log_e(T_i - b))}
$$

@@

**hatchR** assumes input data consists of at least two required columns: a date column indicating the date (and optionally time) of each temperature measurement, and a temperature column providing the corresponding temperature measurement (in °C). Other columns may be present, but columns names should not include spaces. Data should follow the format outlined in Table 1.

@@

Here, we demonstrate the utility three functions--`summarize_temp()`, `plot_check_temp()`, and `check_continuous()`--using a simulated year-long data set (`year_sim`). This dataset contains temperature readings taken every thirty minutes, and its structure (dimensions and first six rows) is shown below.

@@

We demonstrate how the `fit_model()` function may be used to create custom parameterizations for species beyond the Salmonids included in `model_table`. To showcase its utility, we provide parameterizations for three warm-water species: Smallmouth Bass (*Micropterus dolomieu*) [@webster1948], Channel Catfish (*Ictalurus punctatus*) [@small2001], and Lake Sturgeon (*Acipenser fulvescens*) [@smith2005]. These species were selected due to their common use in aquaculture and sport fisheries, illustrating the broad applicability of the effective values approach. For concision, we present parameterization for Smallmouth Bass here, while the full implementation details for all species are available in the `paper.Rmd` on the GitHub project repository (<https://github.com/bmait101/hatchR>).

@@

To demonstrate this case study, we use the **hatchR** graphical user interface available at <https://elifelts.shinyapps.io/hatchR_demo/>. Users begin by uploading their temperature dataset through the `Import Data` window, selecting their file, specifying the appropriate temperature and date columns, and providing the date format (e.g., year-month-day or day-month-year). For this example, we use the `crooked_river` dataset, which is included in **hatchR**.

@@

Following model selection, **hatchR** outputs results in two key locations: the `Phenology Summaries` tab, which provides a table with predictions for each spawn date, and the `Timeline Plot` tab, which shows the corresponding visualization of development timing. Both the prediction table and plot can be downloaded directly from their respective tabs. The full process is demonstrated through a detailed walkthrough on the **hatchR** Shiny website.

@@

We use the `idaho` dataset from @isaak2018, which includes four years of temperature data (2011-2014) for these sites. To identify putative Bull Trout Spawning locations, we apply a filtering criterion based on mean August temperature, as outlined in @isaak2015, selecting only sites with mean August temperature ≤ 13 °C, a known thermal threshold for Bull Trout spawning suitability. The filtering process reduced the dataset to 139 potential spawning sites.

@@

**hatchR** is a software ecosystem that bridges the analytical gap in predicting developmental phenology for wild fishes. It establishes a formal framework for applying effective value models from user-provided parameterizations. The software is available in two formats: 1) A fully customizable R package, ideal for complex and repetitive analyses and 2) a graphical user interface for ease of use, designed for tasks that may only need to be run once or a few times.

@@

Finally, while **hatchR** was designed specifically for fishes, it has potential applications for other poikilothermic organisms, such as reptiles, amphibians, and invertebrates, where developmental rates similarly follow a power law relationship with temperature. Extending the effective value framework to these taxa could provide valuable insights into their life history timing under variable environmental conditions. We encourage users to submit fitted models using this framework as a pull request to **hatchR**'s Github repository provided that data used in the models are published.

@@

Importantly, hatchR extends the effective value framework developed by @sparks2019 into a generalizable tool that can be applied to any fish species or population, provided that appropriate source data are available—data that can easily be collected in aquaculture settings. We present foundational applications of **hatchR**, with additional user-friendly cases and implementation guides available on the software's website. The software is designed for both applied and fundamental research, allowing users to engage with it either through a programmatic R environment or via a user-friendly Shiny app. We expect that the examples provided here represent only a fraction of **hatchR**'s potential applications and encourage the user community to explore and expand upon this framework for their own research and management needs.
